---
title: "aftermath_03"
format: html
editor: visual
---

## Aftermath to exercise 03_gnss

### Step 1: Install Required Libraries

You’ll need some libraries in R for handling CSV files, spatial data, and interactive plotting. You can install them using:

```{r}
#install.packages(c("ggplot2", "leaflet", "dplyr", "readr", "sf"))

```

### Step 2: Load and Prepare Your Data

First, load your data from the CSV files. You can use the `readr` package to read CSV files.

```{r}
library(readr)
library(dplyr)

# Assuming your CSV files are in a directory called "data"
# You can read multiple CSV files into a list and combine them
file_list <- list.files(path = "/Users/maximilianfabi/Desktop/_repos/bsc_2024_25_teaching_inventur_und_geomatik/04_accuracy_inventory/data/csv-files", pattern = "*.csv", full.names = TRUE)

# Read and combine all the CSV files
data <- lapply(file_list, read_csv) %>%
  bind_rows()

# Preview the combined data
head(data)

```

### Step 3: Preprocess Data

Make sure the data is in the right format, especially converting the "Description" (sky cover) and "Title" (points) columns:

```{r}
# Convert the cover to a percentage
data <- data %>%
  mutate(sky_cover = as.numeric(Description) * 10)  # Convert to percentage

# Remove rows with missing or invalid latitude/longitude
data <- data %>%
  filter(!is.na(Latitude), !is.na(Longitude))

# Calculate the average sky cover for each point (group by Title)
data <- data %>%
  group_by(Title) %>%
  mutate(avg_sky_cover = mean(sky_cover, na.rm = TRUE)) %>%
  ungroup()

# Preview the cleaned data
head(data)

```

### Step 4: Visualize with ggplot2

Now you can create a plot using `ggplot2` to show how GPS precision varies based on sky cover. For example, you might use a scatter plot where each point is colored by the level of sky cover:

```{r}
library(ggplot2)

ggplot(data, aes(x = Longitude, y = Latitude, color = sky_cover)) +
  geom_point(size = 3) +
  scale_color_gradient(low = "blue", high = "red") +  # Color scale for cover
  theme_minimal() +
  labs(title = "GPS Precision vs Sky cover",
       color = "Sky cover (%)",
       x = "Longitude", y = "Latitude")

```

### Step 5: Add Interactive Map with `leaflet`

For an interactive experience, you can use `leaflet` to plot the points on a map. This will allow zooming and panning to explore the GPS points.

```{r}
library(leaflet)

leaflet(data) %>%
  addTiles(options = tileOptions(maxZoom = 22)) %>%
  addCircleMarkers(~Longitude, ~Latitude, 
                   color = ~colorNumeric("YlOrRd", sky_cover)(sky_cover),
                   popup = ~paste("Title:", Title, "<br>", "Sky cover:", sky_cover, "%")) %>%
  addLegend("bottomright", pal = colorNumeric("YlOrRd", data$sky_cover),
            values = data$sky_cover,
            title = "Sky cover (%)",
            opacity = 1)

```

```{r}
library(leaflet)
library(RColorBrewer)

# Generate a color palette for each unique Title
color_palette <- colorFactor(palette = brewer.pal(n = length(unique(data$Title)), "Set1"), domain = data$Title)

# Create the leaflet map
leaflet(data) %>%
  addTiles(options = tileOptions(maxZoom = 22)) %>%
  
  # Add circle markers, colored by Title
  addCircleMarkers(
    ~Longitude, ~Latitude,
    color = ~color_palette(Title),
    popup = ~paste("Title:", Title),
    radius = 6
  ) %>%
  
  # Add a legend for the Title colors
  addLegend(
    "bottomright",
    pal = color_palette,
    values = data$Title,
    title = "Point name",
    opacity = 1
  )

```

```{r}
library(geosphere)  # For calculating distances in meters

# Calculate the centroid (mean latitude and longitude) for each point (Title)
centroid_data <- data %>%
  group_by(Title) %>%
  summarise(
    centroid_lat = mean(Latitude),
    centroid_lon = mean(Longitude)
  )

# Directly calculate the distance from each point to its centroid using centroid data
data <- data %>%
  rowwise() %>%
  mutate(
    # Find the corresponding centroid for this point
    centroid_lat = centroid_data$centroid_lat[centroid_data$Title == Title],
    centroid_lon = centroid_data$centroid_lon[centroid_data$Title == Title],
    # Calculate distance to the centroid
    distance_to_centroid = distGeo(c(Longitude, Latitude), c(centroid_lon, centroid_lat))  # Distance in meters
  )

# Calculate variance of distance from centroid for each point (in meters) and include sky cover
variance_data <- data %>%
  group_by(Title, avg_sky_cover) %>%  # Include Description (sky cover)
  summarise(
    variance_distance = var(distance_to_centroid),  # Variance of distance in meters
    avg_distance = mean(distance_to_centroid),  # Mean distance to centroid
    sky_cover = first(Description)  # Keep the sky cover value for each point
  )

# Preview the variance data
print(variance_data)

# Plot: Sky cover vs Variance in GPS Precision
ggplot(variance_data, aes(x = avg_sky_cover, y = avg_distance)) +
  geom_point(color = "blue", size = 3) +  # Plot points for variance
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a trendline
  geom_text(aes(label = Title), hjust = 2, vjust = 2, size = 3) +  # Add point names
  theme_minimal() +
  labs(title = "Average distance to centroid in GPS Precision vs Average Sky cover",
       x = "Sky cover (in %)",
       y = "Average distance (meters) from centroid")
```
